Higher-Order unification for free!

### 2024-04-16 paper/main.tex
still rough 5.18

In order to specify unification, we need to
define the equational theory and
substitution (unification-variable assignment).

@@

\noindent
Proposition \ref{prop:w-preservation} is key to proving \cref{prop:simulation,prop:fidelity}.
Informally, it says that the problematic terms moved on the side by the compiler
are not reintroduced by \hstep, hence \Ue{} can continue to operate properly.
In \cref{sec:simulation,sec:eta,sec:beta}
we describe how the compiler recognizes terms in \maybebeta, \maybeeta and
\notllambda and how \progress takes care of them preserving \wellb
and ensuring \cref{prop:simulation,prop:fidelity,prop:fidelity-recovery}.

@@

The simple compiler described in this section serves as a base for the
extensions in \cref{sec:llam,sec:eta,sec:beta}.
Its main task is to beta normalize the term and map one syntax tree to
the other.
In order to bring back the substitution from \Ho{} to \Fo{} the compiler
builds a ``memory map'' connecting the \Fo{} variables to the \Ho ones
using the routine
\ref{clause:malloc}.

@@

In order to make \Uo higher-order, we need to
take care of terms in \maybebeta.
In the example below, we can see that the basic compilation given in 
the previous section is not able to make the \Ho unification problem succeeds.
%
% make test ONLY=7006 TEX=tex
\printAlllSingle
  {{{\lambda x.(f\appsep (X\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{\lambda x.(f\appsep (A\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{X,A,0}}}
  {{}}
The unification problem $\hoUnifPb_1$ fails while trying to unify
$A \appsep x$ and $x$, which is equivalent to \elpiIn{app [uva A [], x]}
versus \elpiIn{x}.
In order to exploit the higher-order unification algorithm of the meta language,
we need to compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$,
which is \elpiIn{uva A [x]}.

@@

\begin{proof}[Proof sketch]
  We want to prove that for any $\fstep_i$, $\fstep_i$ succeeds iff $\hstep_i$
  succeeds. An $\hstep_i$ is made by a unification $\mathcal{U}$ and a link
  progression $\mathcal{P}$.
  Two cases should be analyzed: 1) $\fstep_i$ succeeds, this means that $\mathcal{U}$
  succeeds since any \maybeeta term is replaced with a fresh variable, and by
  \cref{th:fidelity-beta}, the other subterms unify. Moreover, $\mathcal{P}$
  succeeds, since the $\eta$-progression
  rules simply imitate the rules~\ref{clause:eta1} and \ref{clause:eta2},
  performing unification between $\eta$-contracted terms. 
  If 2) $\fstep_i$ fails, either the failure is 
  due to the unification of terms that are not in \maybeeta. In this case
  $\mathcal{U}$ fails thanks to \cref{th:fidelity-beta}. 
  On the other hand, the failure is
  due to a unification over \maybeeta terms, making $\mathcal{P}$ to fail.
\end{proof}

@@

We give an example for rule
\progressetaright in the next section.

>>>>

% In order to specify unification, we need to
% define the equational theory and
% substitution (unification-variable assignment).

@@

\noindent
Proposition \ref{prop:w-preservation} is key to proving \cref{prop:simulation,prop:fidelity}.
Informally, it says that the problematic terms moved on the side by the compiler
are not reintroduced by \hstep, hence \Ue{} can continue to operate properly.
In \cref{sec:simulation,sec:eta,sec:beta}
we describe how to %the compiler
recognize terms in \maybebeta, \maybeeta and
\notllambda and how \progress takes care of them preserving \wellb
and ensuring \cref{prop:simulation,prop:fidelity,prop:fidelity-recovery}.

@@

The simple compiler described in this section serves as a base for the
extensions in \cref{sec:llam,sec:eta,sec:beta}.
Its main task is to beta normalize the term and map one syntax tree to
the other.
In order to bring back the substitution from \Ho{} to \Fo{} the compiler
builds a ``memory map'' connecting the \Fo{} variables to the \Ho ones
using %the routine
\ref{clause:malloc}.

@@

In order to make \Uo higher-order, we need to
take care of terms in \maybebeta.
In the example below, we can see that the basic compilation given in 
the previous section is not able to make the \Ho unification problem succeeds.
%
% make test ONLY=7006 TEX=tex
\printAlllSingle
  {{{\lambda x.(f\appsep (X\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{\lambda x.(f\appsep (A\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{X,A,0}}}
  {{}}
The unification problem $\hoUnifPb_1$ fails while trying to unify
$A \appsep x$ and $x$, which is equivalent to <<\elpiIn{app [uva A [], x]}>>
versus \elpiIn{x}.
In order to exploit the higher-order unification algorithm of the meta language,
we %need to 
compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$.
%,which is \elpiIn{uva A [x]}.

@@

\begin{proof}[Proof sketch]
  We want to prove that for any $i$, $\fstep_i$ succeeds iff $\hstep_i$
  succeeds. An $\hstep_i$ is made by a unification
  $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  The interesting case is when the infolved terms are in \maybeeta.
  wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
  we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
  In all cases $X \Ue t2$ succeeds since X is a variable and
  hence assigns t2 to X.
  Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
  additional care of eta expanding t2 if needed (5.9.1).

  Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
  $s1$ with $s2$. The interesting case is when fostep succeds by
  finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
  rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
  and that is in exactly the cases in which T (from page 3) is
  not a lam (see 5.9.1) compensating for the missing rules in \Ue.

  2) If $\fstep_i$ fails to unify $s1 \Uo s2$
  it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
  and the one flam-flam we find different terms.
  (5.9.1) only adds lambdas that are forcibly consumed by
  lam-lam rule.
  \end{proof}

Note that \progressetadedup always unifies terms starting with exaclty one
lambda, hence lam-lma, so we can't fail due to that lambda or succed thanks
to that lambda. Hence it does not impact fidelity.

\begin{corollary}[Properties of \Uo in $\wellb_{\beta\eta}$]\label{prop:unifetawb}
  Given the updated compilation scheme, the \Uo of~\cref{sec:founif}
  is a good unification (as per \cref{prop:good-unif})
  for \Eo in the domain $\wellb_{\beta\eta}$.
\end{corollary}

Note \progressetaright is only needed in the next section where chains of links
are added. Hence we give an example for rule \progressetaright in the next section.
\todo{fix english}

@@
